### 2024-07-08 latex/HP3_comm.tex
new input files

%%%%%%%%%%
\subsection*{3.4}
What is ``the empty pathway"?

\begin{translation}[hp03_004]
Suṣumṇā, the empty pathway, the great path to the aperture of Brahmā, the cremation ground, Śāmbhavī, and the middle path are synonyms.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.10}
\begin{translation}[hp03_010]
Now the great seal (\emph{mahāmudrā}):\\
{}[The yogi] should press the perineum with the heel of the left foot, hold [the foot of] the extended leg with the hands and breathe in through the mouth.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.19}
\begin{translation}[hp03_019]
Now, the great lock (\emph{mahābandha}):\\
{}[The yogi] should place the heel of the left foot on the perineal region. And he should put the right foot on the left thigh, [\dots]
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.25}
\begin{translation}[hp03_025]
Now the great piercing (\emph{mahāvedha}):\\
Like a beautiful and charming woman without a man, the great seal and the great lock are barren without the great piercing.\\
\end{translation}

@@

@@

%%%%%%%%%%
\subsection*{3.33}
\begin{translation}[hp03_033]
Now, the sky-roving [seal] (\emph{khecarī}):\\
By cutting, moving, and milking, the yogi should gradually lengthen the tongue until it touches the middle of the brows. Then there is success in \emph{khecarī}.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.50}
\begin{translation}[hp03_050]
The root lock (\emph{mūlabandha})—\\
When [the yogi] presses the perineum with part of the heel, clenches the anus and draws up \emph{apāna}, it is called the root lock.
\end{translation}
% JB: the back part of the heel presses the perineum. It would be better to say, 'with part of the heel.'

@@

%%%%%%%%%%
\subsection*{3.59}
\begin{translation}[hp03_059]
Now the Uḍḍīyaṇa lock:\\
Because the breath flies up (\emph{uḍḍīyate}) into the Suṣumṇā when it is bound by it, yogis say that this [practice] is called Uḍḍīyaṇa.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.67}
\begin{translation}[hp03_067]
Now the \sl{jālandhara} lock–\\
{}[The yogi] should contract the throat and firmly place the chin on the chest. This is the lock called \sl{jālandhara}. It prevents loss of the nectar of immortality.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.73*1}
\begin{translation}[hp03_073_1]
Now the inverted bodily position–\\
The sun devours whatever nectar flows from the divine moon. As a result, the body is afflicted by old age.
\end{translation}
%3.75-76 gray-scaled.

@@

%%%%%%%%%%
\subsection*{3.79}
\begin{translation}[hp03_079]
Now \emph{vajrolī}:\\
Even if he behaves as he wishes without [following] the observances (\emph{niyama}) taught in yoga, the [yogi] who knows \emph{vajrolī} is worthy of success.
\end{translation}

@@

@@

@@

%%%%%%%%%%
\subsection*{3.94*2}
\begin{translation}[hp03_094_2]
Now the Stimulation of the Goddess (\emph{śakticālanam}):\\ 
She whose body is bent (\emph{kuṭilāṅgī}), she who is coiled (\emph{kuṇḍalinī}), the female snake (\emph{bhujaṅgī}), the power (\emph{śakti}), the goddess (\emph{īśvarī}), she who is coiled (\emph{kuṇḍalī}) and Arundhatī: these words are synonyms.
\end{translation}
% 
%\begin{sources}[hp03_094_2]
%\end{sources}

@@

>>>>

%%%%%%%%%%
\subsection*{3.4 heading}
\begin{translation}[hp03_004a]
What is ``the empty pathway"?
\end{translation}

% \begin{philcomm}[hp03_004a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.4}
\begin{translation}[hp03_004]
Suṣumṇā, the empty pathway, the great path to the aperture of Brahmā, the cremation ground, Śāmbhavī, and the middle path are synonyms.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.10 heading}
\begin{translation}[hp03_010a]
Now the great seal (\emph{mahāmudrā}):
\end{translation}

% \begin{philcomm}[hp03_010a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.10}
\begin{translation}[hp03_010]
{}[The yogi] should press the perineum with the heel of the left foot, hold [the foot of] the extended leg with the hands and breathe in through the mouth.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.19 heading}
\begin{translation}[hp03_019a]
Now, the great lock (\emph{mahābandha}):
\end{translation}

% \begin{philcomm}[hp03_019a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.19}
\begin{translation}[hp03_019]
{}[The yogi] should place the heel of the left foot on the perineal region. And he should put the right foot on the left thigh, [\dots]
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.25}
\begin{translation}[hp03_025]
Like a beautiful and charming woman without a man, the great seal and the great lock are barren without the great piercing.\\
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.26 heading}
\begin{translation}[hp03_026a]
Now the great piercing (\emph{mahāvedha}):
\end{translation}

% \begin{philcomm}[hp03_026a]
% \end{philcomm}

@@

%%%%%%%%%%
\subsection*{3.33 heading}
\begin{translation}[hp03_033a]
Now, the sky-roving [seal] (\emph{khecarī}):
\end{translation}

% \begin{philcomm}[hp03_033a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.33}
\begin{translation}[hp03_033]
By cutting, moving, and milking, the yogi should gradually lengthen the tongue until it touches the middle of the brows. Then there is success in \emph{khecarī}.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.50 heading}
\begin{translation}[hp03_050a]
The root lock (\emph{mūlabandha}):
\end{translation}

% \begin{philcomm}[hp03_050a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.50}
\begin{translation}[hp03_050]
When [the yogi] presses the perineum with part of the heel, clenches the anus and draws up \emph{apāna}, it is called the root lock.
\end{translation}
% JB: the back part of the heel presses the perineum. It would be better to say, 'with part of the heel.'

@@

%%%%%%%%%%
\subsection*{3.59 heading}
\begin{translation}[hp03_059a]
Now the Uḍḍīyaṇa lock:
\end{translation}

% \begin{philcomm}[hp03_059a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.59}
\begin{translation}[hp03_059]
Because the breath flies up (\emph{uḍḍīyate}) into the Suṣumṇā when it is bound by it, yogis say that this [practice] is called Uḍḍīyaṇa.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.67 heading}
\begin{translation}[hp03_067a]
Now the \sl{jālandhara} lock:
\end{translation}

% \begin{philcomm}[hp03_067a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.67}
\begin{translation}[hp03_067]
{}[The yogi] should contract the throat and firmly place the chin on the chest. This is the lock called \sl{jālandhara}. It prevents loss of the nectar of immortality.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.74 heading}
\begin{translation}[hp03_074a]
Now the inverted bodily position:
\end{translation}

% \begin{philcomm}[hp03_074a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.73*1}
\begin{translation}[hp03_073_1]
The sun devours whatever nectar flows from the divine moon. As a result, the body is afflicted by old age.
\end{translation}
%3.75-76 gray-scaled.

@@

%%%%%%%%%%
\subsection*{3.79 heading}
\begin{translation}[hp03_079a]
Now \emph{vajrolī}:
\end{translation}

% \begin{philcomm}[hp03_079a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.79}
\begin{translation}[hp03_079]
Even if he behaves as he wishes without [following] the observances (\emph{niyama}) taught in yoga, the [yogi] who knows \emph{vajrolī} is worthy of success.
\end{translation}

@@

%%%%%%%%%%
\subsection*{3.88 heading}
\begin{translation}[hp03_088a]
Now \emph{sahajolī}:
\end{translation}
% atha sahajolī/
% \begin{philcomm}[hp03_088a]
% \end{philcomm}

@@

%%%%%%%%%%
\subsection*{3.91 heading}
\begin{translation}[hp03_091a]
Now \emph{amarolī}:
\end{translation}

% \begin{philcomm}[hp03_091a]
% \end{philcomm}

@@

%%%%%%%%%%
\subsection*{3.94*2 heading}
\begin{translation}[hp03_094_2a]
Now the Stimulation of the Goddess (\emph{śakticālanam}):
\end{translation}

% \begin{philcomm}[hp03_094_2a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{3.94*2}
\begin{translation}[hp03_094_2]
She whose body is bent (\emph{kuṭilāṅgī}), she who is coiled (\emph{kuṇḍalinī}), the female snake (\emph{bhujaṅgī}), the power (\emph{śakti}), the goddess (\emph{īśvarī}), she who is coiled (\emph{kuṇḍalī}) and Arundhatī: these words are synonyms.
\end{translation}
% 
%\begin{sources}[hp03_094_2]
%\end{sources}

@@

%%%%%%%%%%
\subsection*{3.103*1 ending}
% iti śakticālanaṃ/
\begin{translation}[hp03_103_1p]
\end{translation}

% \begin{philcomm}[hp03_103_1p]
% \end{philcomm}

### 2024-07-08 latex/HP4_comm.tex
new input files

@@

%%%%%%%%%%
\subsection*{4.8*2}
\begin{translation}[hp04_008_2]
Some are confused by a multitude of [tantric] texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.8*8}
\begin{translation}[hp04_008_8]
Now \emph{khecarī}:\\
There is a hollow that generates knowledge and has five streams. \emph{Khecarīmudrā} is situated in that pure void.
\end{translation}
%
%\begin{sources}[hp04_008_8]
%\end{sources}

@@

%%%%%%%%%%
\subsection*{4.8*22}
\begin{translation}[hp04_008_22]
Now \emph{śāmbhavī}:\\
{[}The yogi] should [put] the mind in Śakti and Śakti in the mind, observe the mind with the mind, and meditate on it as the supreme state.
\end{translation}
%V3 is closer to the CA reading

@@

@@

%%%%%%%%%%
\subsection*{4.14}
\begin{translation}[hp04_014]
What kind of indifference?\\
In the cold season, [indifference towards] whether [one has] an open tent or a quilt, with regard to good nourishment, whether it is cow's milk or water, with regard to food, whether it is lots of alms or forest roots, and with regard to the vessel for food, whether it is the hand or some kind of bucket.
%
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.17}
\begin{translation}[hp04_017]
Among these, the beginning stage is [as follows]:\\
As a result of the piercing of the knot of Brahmā, bliss arises in the void [and] an unstruck sound which has various tones is heard in the body.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.19}
\begin{translation}[hp04_019]
Now the unified stage:\\
In the second stage the breath, after bringing about union, goes into the middle [channel]. Then the yogi has a firm posture [and] he becomes a gnostic and equal to a god.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.20}
\begin{translation}[hp04_020]
Then, as a result of the piercing of the knot of Viṣṇu there is a hint of supreme bliss in total emptiness (\emph{atiśūnye}), and then the pounding sound of a \textit{bherī} drum occurs.%
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.21}
\begin{translation}[hp04_021]
Now the accumulation stage:\\
In the third [stage], having pierced [the knot of Viṣṇu], the sound of a \textit{mardala} drum arises in space. Then [the yogi] reaches the great void, the abode of all supernatural powers.%
\end{translation}
%

@@

%%%%%%%%%%
\subsection*{4.23}
\begin{translation}[hp04_023]
Now the perfection stage:\\
Then, having pierced Rudra's knot, the breath goes to all the seats [of the deities in the body]. In the perfected [stage] the sound of a flute becomes the sound of a resonating lute.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.24}
\begin{translation}[hp04_024]
Then the mind having become one [with the sound]. This is called Rājayoga. He becomes a creator and destroyer, an equal to a lord among yogis.%
\end{translation}

@@

\begin{testimonia}[hp04_044]
\emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{purā matsyendrabodhāya ādināthoditavacaḥ |\\+}
\tl{manaḥ pākam avāpnoti nirālambākhyaghoṭanam ||\\!}
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{4.47*1}
\begin{translation}[hp04_047_1]
The inner sound is a net for trapping the deer of the mind and a hunter for corraling the antelope of the mind.
\end{translation}
% nādo’ntaraṅgasāraṅgabandhane vāgurāyate
%V15 antaraṅgaturaṅgasya nādo vyādhāyate pi ca

@@

@@

\begin{metre}[hp04_055]
Anuṣṭubh (a: ma-vipulā)
\end{metre}
%%%%%%%%%%
\subsection*{4.56}
\begin{translation}[hp04_056]
And so, Vasiṣṭha [said]:\\
The moon and sun move in Iḍā and Piṅgalā. The moon is said to be of the nature of \emph{tamas} and the sun of \emph{rajas}.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.58}
\begin{translation}[hp04_058]
For as the tetrad of verses called the Saubhadra has it:\\
There are six cakras, sixteen supports, three focal points and three \emph{guṇa}s. Everything else is [just] the prolixity of texts. Trikūṭa is the supreme place.
\end{translation}

@@

\begin{philcomm}[hp04_058]
We do not know why this tetrad of verses is called Saubhadra.\\

The three components of the yogic body listed here are found together in other texts, the earliest being \emph{Netratantra} 7.1ab (\emph{ṛtucakraṃ svarādhāraṃ trilakṣyaṃ vyomapañcakam}). However, we are yet to find a source for this list that includes the three \emph{guṇa}s.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.70}
\begin{translation}[hp04_070]
Thus, Viśvarūpācārya [said]:\\
When the breath is destroyed and the mind dissolves, all experience is the same (\emph{samarasatva}). That is called \emph{samādhi}.
\end{translation}

>>>>

%%%%%%%%%%
\subsection*{4.1 heading}
\begin{translation}[hp04_001a]
Now \emph{samādhi}:
\end{translation}

% \begin{philcomm}[hp04_001a]
% \end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.8*2}
\begin{translation}[hp04_008_2]
Some are confused by a multitude of tantric texts, some by a mass of vedic texts and some by reasoning. They do not know what causes one to cross over (\emph{tārakam}).%
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.8*8 heading}
\begin{translation}[hp04_008_8a]
Now \emph{khecarī}:
\end{translation}

% \begin{philcomm}[hp04_008_8a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.8*8}
\begin{translation}[hp04_008_8]
There is a hollow that generates knowledge and has five streams. \emph{Khecarīmudrā} is situated in that pure void.
\end{translation}
%
%\begin{sources}[hp04_008_8]
%\end{sources}

@@

%%%%%%%%%%
% \subsection*{4.8*21 ending}
% iti khecarī/
% \begin{translation}[hp04_008_21p]
% \end{translation}

% \begin{philcomm}[hp04_008_21p]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.8*22 heading}
\begin{translation}[hp04_008_22a]
Now \emph{śāmbhavī}:
\end{translation}

% \begin{philcomm}[hp04_008_22a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.8*22}
\begin{translation}[hp04_008_22]
{[}The yogi] should [put] the mind in Śakti and Śakti in the mind, observe the mind with the mind, and meditate on it as the supreme state.
\end{translation}
%V3 is closer to the CA reading

@@

%%%%%%%%%%
% \subsection*{4.33_8 heading}
% atha viśrāntiḥ/
% \begin{translation}[hp04_033_8a]
% \end{translation}

% \begin{philcomm}[hp04_033_8a]
% \end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.14 heading}
\begin{translation}[hp04_014a]
What kind of indifference?
\end{translation}

% \begin{philcomm}[hp04_014a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.14}
\begin{translation}[hp04_014]
In the cold season, [indifference towards] whether [one has] an open tent or a quilt, with regard to good nourishment, whether it is cow's milk or water, with regard to food, whether it is lots of alms or forest roots, and with regard to the vessel for food, whether it is the hand or some kind of bucket.
%
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.17 heading}
\begin{translation}[hp04_017a]
Among these, the beginning stage is [as follows]:
\end{translation}

% \begin{philcomm}[hp04_017a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.17}
\begin{translation}[hp04_017]
As a result of the piercing of the knot of Brahmā, bliss arises in the void [and] an unstruck sound which has various tones is heard in the body.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.19 heading}
\begin{translation}[hp04_019a]
Now the unified stage:
\end{translation}

% \begin{philcomm}[hp04_019a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.19}
\begin{translation}[hp04_019]
In the second stage the breath, after bringing about union, goes into the middle [channel]. Then the yogi has a firm posture [and] he becomes a gnostic and equal to a god.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.20}
\begin{translation}[hp04_020]
Then, as a result of the piercing of the knot of Viṣṇu there is a hint of supreme bliss in total emptiness (\emph{atiśūnye}), and then the pounding sound of a kettle drum occurs.%
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.21 heading}
\begin{translation}[hp04_021a]
Now the accumulation stage:
\end{translation}

% \begin{philcomm}[hp04_021a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.21}
\begin{translation}[hp04_021]
In the third [stage], having pierced [the knot of Viṣṇu], the sound of a bass drum arises in space. Then [the yogi] reaches the great void, the abode of all supernatural powers.%
\end{translation}
%

@@

%%%%%%%%%%
\subsection*{4.23 heading}
\begin{translation}[hp04_023a]
Now the perfection stage:
\end{translation}

% \begin{philcomm}[hp04_023a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.23}
\begin{translation}[hp04_023]
Then, having pierced Rudra's knot, the breath goes to all the seats [of the deities in the body]. In the perfected [stage] the sound of a flute becomes the sound of a resonating lute.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.24}
\begin{translation}[hp04_024]
Then the mind becomes one [with the sound]. This is called Rājayoga. He becomes a creator and destroyer, an equal to a lord among yogis.%
\end{translation}

@@

\begin{testimonia}[hp04_044]
\emph{Yogacintāmaṇi} f.~26v (attr. to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{purā matsyendrabodhāya ādināthoditavacaḥ |\\+}
\tl{manaḥpākam avāpnoti nirālambākhyaghoṭanam ||\\!}
\end{versinnote}

@@

%%%%%%%%%%
\subsection*{4.47*1}
\begin{translation}[hp04_047_1]
The inner sound is a net for trapping the deer of the mind and a hunter for corraling the antelope of the mind.
\end{translation}
% ?? JB: we have not translated antaraṅga° in pāda a. Change to 'trapping the inner deer that is the mind'?

@@

%%%%%%%%%%
\subsection*{4.52*4 ending}
% iti nādānusaṃdhānam/
\begin{translation}[hp04_052_4p]
\end{translation}

% \begin{philcomm}[hp04_052_4p]
% \end{philcomm}

@@

\begin{metre}[hp04_055]
Anuṣṭubh (a: ma-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{4.56 heading}
\begin{translation}[hp04_056a]
And so, Vasiṣṭha [said]:
\end{translation}

% \begin{philcomm}[hp04_056a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.56}
\begin{translation}[hp04_056]
The moon and sun move in Iḍā and Piṅgalā. The moon is said to be of the nature of \emph{tamas} and the sun of \emph{rajas}.
\end{translation}

@@

%%%%%%%%%%
\subsection*{4.58 heading}
\begin{translation}[hp04_058a]
For as the tetrad of verses called the Saubhadra has it:
\end{translation}

\begin{philcomm}[hp04_058a]
We do not know why this tetrad of verses is called Saubhadra.
\end{philcomm}

%%%%%%%%%%
\subsection*{4.58}
\begin{translation}[hp04_058]
There are six cakras, sixteen supports, three focal points and three \emph{guṇa}s. Everything else is [just] the prolixity of texts. Trikūṭa is the supreme place.
\end{translation}

@@

\begin{philcomm}[hp04_058]
The three components of the yogic body listed here are found together in other texts, the earliest being \emph{Netratantra} 7.1ab (\emph{ṛtucakraṃ svarādhāraṃ trilakṣyaṃ vyomapañcakam}). However, we are yet to find a source for this list that includes the three \emph{guṇa}s.
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{4.70 heading}
\begin{translation}[hp04_070a]
Thus, Viśvarūpācārya [said]:
\end{translation}

% \begin{philcomm}[hp04_070a]
% \end{philcomm}

%%%%%%%%%%
\subsection*{4.70}
\begin{translation}[hp04_070]
When the breath is destroyed and the mind dissolves, all experience is the same (\emph{samarasatva}). That is called \emph{samādhi}.
\end{translation}
